---
# global document parameters
title: "Exercise 1: Presidential Polls in 2012"
author: "PS 270: Understanding Political Numbers"
date: "Due Wednesday, February 27"
# abstract: 
# bibliography: /Users/michaeldecrescenzo/Dropbox/bib.bib
# biblio-style: authoryear

# specification of output
# we use bookdown::pdf_document2 to cross-reference tables and figs
output: 
  bookdown::pdf_document2: 
    toc: false # true default?
    keep_tex: true
    includes: 
      in_header: rmd-preamble.tex
    number_sections: false
    highlight: tango
    fig_caption: true
    citation_package: natbib

# can control some latex-type things with the YAML front matter
# it may be less cluttered if you control some of these options 
#   in a preamble or template file instead
fontsize: 12pt
geometry: margin = 1.25in
indent: false
linkcolor: red 
urlcolor: blue 
citecolor: black
subparagraph: yes
---

```{r, echo = FALSE}
# chunks:
# hide code and messages
# cache everything
knitr::opts_chunk$set(include = TRUE, eval = FALSE, echo = TRUE,
                      warning = FALSE, message = FALSE,
                      cache = TRUE, 
                      fig.align = "center", # eval.after = 'fig.cap',
                      fig.retina = 3
                      # , dpi = 100
                      )
```

<!-- # Introduction -->

In this assignment, we will investigate polling data from three swing states in the 2012 presidential election: Florida, Ohio, and Virginia. Each case is a poll showing the support for Barack Obama (D) and Mitt Romney (R).

This exercise will test some important R skills for interacting with data. We will import data into R, examine and modify the data, and conduct simple analysis with a graphic.

**Before beginning the assignment**, do the following steps to prepare your computer.

1. On your computer, you should have a folder titled `PS-270` (or something similar) that contains all of your work for this class. *Inside* of your class folder, create two more folders (if you haven't already): one called `data` and another called `figures`.

2. Download and save the polling data from Canvas into your new `data` folder.

3. Open RStudio by double-clicking your `.Rproj` file inside your `PS-270` folder.

4. Run the following commands in the console to make sure that everything is set up to begin the assignment.
    a. If you enter the `getwd()` function, R should print a file pathway that ends with the name of your `PS-270` folder.
    b. The `list.files()` function prints the names of files and folders in your `PS-270` folder. You should see your `"data"` and `"figures"` folders among the output.
    c. Typing `list.files("data")` prints the names of files and folders inside the `data` folder. You should find the `.csv` data that you downloaded from Canvas.


# Your Tasks

Write and save an R script that implements the following tasks in order.

```{r, echo = FALSE}
library("tidyverse")
library("here")
```

1. Load the `tidyverse` and `here` packages.^[
  Install any missing packages by typing `install.packages("package_name")` into the console. Don't save any package installation commands in your R script file.
]

2. Import the polling data file into R. Do this with the `read_csv()` function as follows.
    ```{r}
polls <- read_csv(here("data", "FileNameGoesHere.csv")) 
    ```
    This code says "import a CSV (spreadsheet) file, which is located in the `data` folder, named `FileNameGoesHere.csv`." You will need to fix the file name so that R grabs the file you downloaded from Canvas.

3. Print the data object to the console to see what's inside.^[
  Tip: You can type `View(dataset_name)` into the console to see a spreadsheet view of the data.
]
Print the variable names.

4. How many polls were taken in each state? Find out by using the `count()` function to tabulate the state variable.

5. Obama's "margin" measures how much more support he had in the poll compared to Romney. It will be a negative number if Romney did better in a given poll. Create a histogram of Obama's vote margin. Plot states as separate panels (using one of the `facet_` functions). Use your judgment and taste to make your graphic look professional: polish up any sloppy axis titles, add a title, and so on. Save the graphic into your `figures` folder.
    ```{r}
ggsave(here("figures", "ex-1_histogram.pdf"))
    ```

6. We often measure U.S. voting using a political party's *share of the two-party vote*: the vote percentage for a party when we set aside Independent and third-party voters. We would calculate the *Democratic share of the two-party vote* using the following equation.
\begin{align*}
  \text{Dem.\ share of two-party vote} &= \frac{\text{Dem.\ vote}}{\left(\text{Dem.\ vote} + \text{Rep.\ vote}\right)}
\end{align*}
Create this variable using `mutate()` and call it `dem_2party`.^[
  Tip: don't overwrite the original data object until you are confident that the code is working without any issues!
]

7. Calculate the mean of Obama's two-party vote share *in each state*. You can do this in one of two ways.
    - Use `filter()` to split the data into three separate objects, one for each state. Then use `summarize()` to calculate the mean of the two-party vote in each state.
    - Group the combined data (`polls`) by state, and find the mean in each group. If you `group_by()` the state, calculating the mean with `summarize()` would give you the mean in each group.

When you are finished, upload both your R script and your saved graphic to Canvas as your submission for the "Exercise 1" assignment.


<!-- 

We want to get them comfortable using basic commands in R.

- Running commands
- Saving scripts
- Simple data frame verbs from the tidyverse
- accessing a variable within a data frame


Outline:

- Print a stock-loaded dataset (like `gapminder`)
- `select()` a subset of variables 
- `filter()` a set of rows
- `arrange()` according to some variable
- Print a `df$variable` and maybe do some simple operation with it or something -->