


---

## Geoms

```{r start-ex-line, fig.show = "hide"}
ggplot(data = gapOC,
       mapping = aes(x = year, y = lifeExp)) +
  geom_line() #<<
```

.pull-left[
```{r start-ex-line, out.width = "100%", fig.height = 3, fig.width = 4, echo = FALSE}
```
]

--



.pull-right[
.center[Why does this look weird?]
```{r}
count(gapOC, country)
```
]


???

`geom_line` always tries to draw a line from left-to-right

- if you have multiple groups of data that you want to be plotted as separate lines, R doesn't know until you tell it



---

## More aesthetics

```{r start-aes-color, fig.show = "hide"}
ggplot(data = gapOC,
       mapping = aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country)) #<<
```


.left-code[
**Grammar: aesthetics.** How does data become a plot feature?

Translation: "I want different colors for each country"

Which aesthetics can be modified? Check the help file: `?geom_line` 
]

--

.right-plot[
```{r start-aes-color, out.width = "100%", fig.height = 3, fig.width = 5, echo = FALSE}
```
]



---

## More aesthetics

```{r start-color, fig.show = "hide"}
ggplot(data = gapOC,
       mapping = aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country, linetype = country)) + #<<
  geom_point(color = "gray50", size = 2, shape = 15) #<<
```

.left-code[
- Layering geoms
- Aesthetics that *don't come from your data* don't need `aes()`
- [Color names in R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)
]

.right-plot[
```{r start-color, out.width = "100%", fig.height = 2.5, fig.width = 5, echo = FALSE}
```
]


---

# More Geoms (and Labels)

.pull-left[
```{r hist, fig.show = "hide"}
ggplot(gapminder, aes(x = lifeExp)) + 
  geom_histogram() + 
  labs(x = "Life Expectancy", 
       y = "Frequency", 
       title = "Histogram of Life Expectancy", 
       subtitle = "Data from 1952-2007") 
```

- For distributions (histograms and densities), only declare `x`
- Specify labels using `labs()` (note quotes and commas)

]

.pull-right[
```{r hist, echo = FALSE, fig.width = 4, fig.height = 3.5, out.width = "100%"}
```
]

---

## Facets

```{r facets, fig.width = 6, fig.height = 3.5, out.width = "60%"}
ggplot(gapminder, 
       aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) #<<
```

???

read the tilde as "as a function of"

- wrap lets the facets wrap around lines
- there is also a `facet_grid` function (which we can get into later)

---

## 'Group' aesthetic

```{r grp, fig.width = 6, fig.height = 3.5, out.width = "60%"}
ggplot(gapminder, 
       aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) +
  geom_line(aes(group = country, color = continent), #<<
            show.legend = FALSE) #<<
```

???

- would have made a legend

---

## Scales

```{r scales, fig.width = 6, fig.height = 3.5, out.width = "60%"}
ggplot(gapminder, 
       aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) +
  geom_line(aes(group = country, color = continent),
            show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") + #<<
  scale_x_continuous(breaks = seq(1960, 2000, 20)) #<<
```

???

- scales modify the default aesthetic settings
- default color palette changed to "set2"
- `scale_aesName_modifier`
  - `scale_x_continuous`, `scale_x_discrete`, `scale_x_log10`
- set the axis ticks (create a sequence of values)


---

## Coordinates

```{r coords, fig.width = 6, fig.height = 3.5, out.width = "60%"}
ggplot(gapminder, 
       aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) +
  geom_line(aes(group = country, color = continent),
            show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") + 
  scale_x_continuous(breaks = seq(1960, 2000, 20)) +
  coord_cartesian(ylim = c(20, 100)) #<<
```



---

## Themes!

```{r themes, fig.width = 6, fig.height = 3.5, out.width = "60%"}
ggplot(gapminder, 
       aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) +
  geom_line(aes(group = country, color = continent),
            show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") + 
  scale_x_continuous(breaks = seq(1960, 2000, 20)) +
  coord_cartesian(ylim = c(20, 100)) +
  theme_minimal() #<<
```



---



```{r}
gapAM <- filter(gapminder, continent == "Americas")

ggplot(gapAM, aes(x = year, y = lifeExp)) +
  geom_line(aes(group = country, color = continent),
            show.legend = FALSE) +
  geom_point(data = filter(gapminder, country == "Rwanda"))

ggplot(gapminder, aes(x = year, y = lifeExp)) +
  facet_wrap(~ continent) +
  geom_line(aes(group = country))

ggplot(gapminder, aes(x = continent)) +
  geom_histogram(stat = "count")
```

<!--
ggsave() 
scale_color_manual
facet_grid
coord_flip
Is there a good online guide?
 -->