---
title: "Nonlinear Relationships"
subtitle: "(Grab \"Essay 2\" Assignment Sheet)"
author: "Understanding Political Numbers"
date: "March 27, 2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    # mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_SVG"
    css: xaringan-themer.css
    nature:
      ratio: "16:9"
      highlightStyle: default
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
seal: false
---


class: center, middle, inverse

# Where do you find nonlinear relationships?

```{r setup-rmd, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}

# rmarkdown::render(here::here("lectures", "17-nonlinear", "17-nonlinear.Rmd"))
# knitr::purl(here::here("lectures", "17-nonlinear", "17-nonlinear.Rmd"), output = here::here("R", "lecture_nonlinear.R"))

source(here::here("R", "setup-lectures.R"))

# They're good DAGs, Brent
library("dagitty")
library("ggdag")

library("broom")

library("patchwork")
library("gapminder")


dblue <- "#259FDD"
rred <- "#FC5E47"

# box
# library("boxr"); box_auth()

# library("viridis")
# library(png)
# library(grid)
# library(gridExtra)

# options(scipen = 99999)


hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})

# chunks:
# hide code and messages
# cache everything
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, 
                      warning = FALSE, message = FALSE,
                      cache = TRUE, 
                      cache.path = here::here("lectures", "cache", "17_"),
                      fig.align = "center", # eval.after = 'fig.cap',
                      fig.retina = 3 # , dpi = 100
                      )

img <- "lectures/17-nonlinear/img"
```



---

## Diminishing Effects

```{r, fig.width= 5, fig.height = 4, out.width = "50%"}
ggplot(midwest, aes(x = percollege, y = percbelowpoverty)) +
  geom_point(shape = 21, color = "white", fill = "steelblue") +
  geom_smooth(se = FALSE, color = "black") +
  labs(x = "Percent College Graduates", y = "Percent Below Poverty Line",
       title = "Poverty and Education",
       subtitle = "Data from Midwest Counties")
```



---

class: middle

```{r, fig.width = 7, fig.height = 3.5, out.width = "80%"}
nanas <- 
  tibble(bananas = 0:5) %>%
  mutate(`Total Utility` = case_when(bananas == 0 ~ 0,
                             bananas > 0 ~ 1.5 + log(bananas)),
         `Marginal Utility` = `Total Utility` - lag(`Total Utility`)) %>%
  gather(key = type, value = value, ends_with("Utility")) %>%
  mutate(type = fct_rev(as.factor(type))) 

ggplot(nanas, aes(x = bananas, y = value)) +
  facet_wrap(~ type, scales = "free") +
  geom_line() +
  geom_point() +
  labs(x = "Number of Bananas", y = "Utiles",
       title = "Value (Utility) of Eating Bananas")
```




---

## Natural Boundaries

```{r, fig.width= 6, fig.height = 4, out.width = "60%"}
ggplot(midwest, aes(x = poptotal, y = percblack)) +
  geom_point(shape = 21, color = "white", fill = "steelblue") +
  scale_x_continuous(trans = "log", labels = function(x) {round(x, -3) %>% scales::comma()}) +
  labs(x = "Population Total", y = "Percent Black/AA",
       title = "Racial Composition and Population",
       subtitle = "Data from Midwest Counties") +
  geom_smooth(color = "black", se = FALSE)
```


---

## "Multiplicative data"

```{r, fig.width= 6, fig.height = 4, out.width = "60%"}
ggplot(midwest, aes(y = percbelowpoverty, x = poptotal)) + 
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  labs(x = "Population", y = "Percent Below Poverty")
```



---

## Long right tails (imperfect indicator)

```{r long-tails, fig.width = 10, fig.height = 5, out.width = "90%"}
h_area <- ggplot(midwest, aes(x = area)) + geom_histogram()
h_poptotal <- ggplot(midwest, aes(x = poptotal)) + geom_histogram()
h_popdensity <- ggplot(midwest, aes(x = popdensity)) + geom_histogram()

h_gappop <- ggplot(gapminder, aes(x = pop)) + geom_histogram()
h_gdp <- ggplot(gapminder, aes(x = gdpPercap)) + geom_histogram()

(h_area + h_poptotal + h_popdensity) / (h_gappop + h_gdp)
```







---

class: inverse, middle, center

## Logarithmic transformations



---

class: middle

```{r, fig.width = 8, fig.height = 4, out.width = "90%"}
log_data <- 
  tibble(x = rnorm(1000, 0, 0.95) %>% exp(), 
         y = log(x) + rnorm(1000, 0, 1))

log_scatter <- ggplot(log_data, aes(x = x, y = y)) + 
  geom_point() +
  labs(title = "Logarithmic relationship",
       subtitle = "Sudden increase that tapers off")

log_hist <- ggplot(data = log_data, aes(x = x)) + 
  geom_histogram() +
  labs(title = "Skewed histogram",
       subtitle = "Long right tail")

log_scatter + log_hist
```


---

class: middle

```{r, fig.width = 8, fig.height = 4, out.width = "90%"}
(log_scatter + 
   aes(x = log(x)) + 
   labs(title = "Logarithmic relationship",
        subtitle = "Appears LINEAR after logging")) +
(log_hist + 
   aes(x = log(x)) + 
   labs(title = "Skewed histogram",
        subtitle = "Appears NORMAL after logging"))
```




---

class: middle

```{r, fig.width = 8, fig.height = 4, out.width = "90%"}
exp_data <- 
  tibble(x = rnorm(1000, 0, 0.95), 
         y = exp(x + rnorm(1000, 0, 1)))

exp_scatter <- ggplot(exp_data, aes(x = x, y = y)) + 
  geom_point() +
  labs(title = "Exponential relationship",
       subtitle = "Explosive nonlinear increase")

exp_scatter_log <- ggplot(exp_data, aes(x = x, y = log(y))) + 
  geom_point() +
  labs(title = "Exponential relationship",
       subtitle = "Appears LINEAR after logging Y")

exp_scatter + exp_scatter_log
```




---

## What is log? (Baby don't hurt me)

.pull-left[
- Steep increase that diminishes

- *Never* fully "levels off"

- Defined over *positive* values only

- Commonly appears with count data, money, population
]

.pull-right[

```{r, fig.width = 4, fig.height = 3, out.width = "100%"}
log_exp <- 
tibble(x = seq(-10, 20, .01),
       logx = log(x),
       ex = exp(x)) %>%
  gather(key = fn, value = y, logx, ex) 

filter(log_exp, fn == "logx") %>%
  ggplot(aes(x = x, y = log(x))) +
    geom_line(linetype = 2) +
    labs(y = "y = log(x)") +
    coord_cartesian(xlim = c(-2, 20))
```
]

---

## Logarithms "undo" exponentials

If $b^{x} = y$, then $\log_b(y) = x$

--

We usually only care about "base $e$" ( $e = `r exp(1)`\ldots$ )

--
.pull-left[
\begin{align}
  e^{x} = y
\end{align}
]

.pull-right[
\begin{align}
  \ln(y) &= x \\
  \log(y) &= x
\end{align}
]

--

Never worry about solving by hand

```{r, echo = TRUE}
# natural log (base e) of 8
log(8)

# exponentials e^(2.079...)
exp(2.079442)
```



---

## Logs and exponentials are *inverse functions*

.pull-left[
Inverse functions: if $y = f(x)$, then $f^{-1}(y) = x$


\begin{align}
  f^{-1}(f(x)) &= x \\[12pt]
  \log(e^{x}) &= x \\[12pt]
  e^{\log(x)} &= x
\end{align}

If you need to log a variable: `log(var)`

If you need to "unlog" a variable: `exp(var)`
]
.pull-right[

```{r, fig.width = 5, fig.height = 4, out.width = "100%"}
  ggplot(log_exp, aes(x = x, y = y)) +
    geom_abline(linetype = 2, color = "gray") +
    geom_hline(yintercept = 0, color = "gray") +
    geom_vline(xintercept = 0, color = "gray") +
    geom_line(aes(group = fn)) +
    coord_cartesian(xlim = c(-2.5, 5),
                    ylim = c(-2.5, 5)) +
    annotate(geom = "text", x = 4, y = 1, label = TeX("$\\log(x)$")) +
    annotate(geom = "text", x = 1.5, y = 3, label = TeX("$e^x$")) +
    annotate(geom = "text", x = 4, y = 4, label = TeX("$y = x$"),
             vjust = -2)

```
]




---

## Why we log for "multiplicative" data

.left-code[
Sum of random fluctuations -> Normal

Product of random fluctuations -> "Log Normal"

```{r, include = FALSE}
clt <- tibble(
  x1 = runif(1000, 1, 10),
  x2 = runif(1000, 1, 10),
  x3 = runif(1000, 1, 10),
  x4 = runif(1000, 1, 10),
  x5 = runif(1000, 1, 10),
  x6 = runif(1000, 1, 10),
  x7 = runif(1000, 1, 10),
  x8 = runif(1000, 1, 10),
  x9 = runif(1000, 1, 10),
  x10 = runif(1000, 1, 10),
  summed = x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9 + x10,
  mult = x1 * x2 * x3 * x4 * x5 * x6 * x7 * x8 * x9 * x10
) %>%
print()
```
]

.right-plot[
```{r, fig.width = 6, fig.height = 3, out.width = "100%"}
normal <- ggplot(clt, aes(x = summed)) + 
  geom_histogram(boundary = 0, binwidth = 1) +
  labs(x = "Sum of random variables")
lognormal <- ggplot(clt, aes(x = mult)) + 
  geom_histogram(boundary = 0) +
  labs(x = "Product of random variables")
normal + lognormal
```
]

--


**Logs turn *multiplicative* operations into *additive* (linear) operations**

\begin{align}
  \log\left(a \times b \right) &= \log\left(a\right) + \log\left(b\right)
\end{align}



---

## Remember those long tails?

```{r long-tails, fig.width = 10, fig.height = 5, out.width = "90%"}
```



---

## Take the log, long tails look normal

```{r log-long-tails, fig.width = 10, fig.height = 5, out.width = "90%"}
h_area <- ggplot(midwest, aes(x = log(area))) + geom_histogram()
h_poptotal <- ggplot(midwest, aes(x = log(poptotal))) + geom_histogram()
h_popdensity <- ggplot(midwest, aes(x = log(popdensity))) + geom_histogram()

h_gappop <- ggplot(gapminder, aes(x = log(pop))) + geom_histogram()
h_gdp <- ggplot(gapminder, aes(x = log(gdpPercap))) + geom_histogram()

(h_area + h_poptotal + h_popdensity) / (h_gappop + h_gdp)

```



---

## Logs in practice

.pull-left[

```{r gap-unlog, echo = TRUE, fig.show = 'hide'}
library("gapminder")

# most recent gapminder year
gap_07 <- gapminder %>%
  filter(year == max(year)) 

# linear relationship looks bad
ggplot(gap_07, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "GDP per cap", y = "Life Exp")
```

]

.pull-right[
```{r gap-unlog, fig.width = 5, fig.height = 4, out.width = "100%"}
```

]

---

## Logs in practice

.pull-left[

Plot $y = f(log(x))$

```{r gap-log, echo = TRUE, fig.show = 'hide'}
# log(x) is... better
ggplot(gap_07, 
       aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "log(GDP per cap)", y = "Life Exp")
```
]


.pull-right[
```{r gap-log, fig.width = 5, fig.height = 4, out.width = "100%"}
```
]



---

## Logs in practice

.pull-left[

Estimate the model

```{r, echo = TRUE}
# new variable is log(x)
gap_07 <- gap_07 %>%
  mutate(log_gdp = log(gdpPercap))

log_model <- lm(lifeExp ~ log_gdp, data = gap_07)

tidy(log_model)
```
]

--

.pull-right[
\begin{align}
  \hat{\mathrm{life}} &= `r coef(log_model)[1] %>% round(2)` + `r coef(log_model)[2] %>% round(2)`\log(\mathrm{gdp.pc})
\end{align}

As *log GDP per capita* increases by one unit...
]




---



.pull-left[
## Logs in practice

Interpret graphically

```{r, echo = TRUE}
# get predicted values for log(x)
# calculate unlogged x
# create upper & lower conf interval bounds
log_preds <- augment(log_model) %>%
  mutate(gdp_pc = exp(log_gdp),
         MOE = 1.96 * .se.fit,
         conf.low = .fitted - MOE,
         conf.high = .fitted + MOE) %>%
  print()
```

]

--

.pull-right[
```{r, echo = TRUE, fig.width = 5, fig.height = 4, out.width = "100%"}
# plot y over unlogged x, add yhat line
ggplot(log_preds, aes(x = gdp_pc, y = lifeExp)) +
  geom_point(color = "gray50") +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high), 
    fill = "maroon", alpha = .3
  ) +
  geom_line(aes(y = .fitted))
```

]




---

## Other log rules

\begin{align}
  \log\left(a \times b \right) &= \log\left(a\right) + \log\left(b\right) \\[12pt]
  \log\left(\frac{a}{b}\right) &= \log\left(a\right) - \log\left(b\right) \\[12pt]
  \log\left(a^{b}\right) &= b \times \log\left(a\right) \\[12pt] 
  \log\left(1\right) &= 0 \\[12pt] 
  \log\left(0\right) &= ?
\end{align}



---

class: center, middle

## Remember...

Diminishing effects, natural boundaries

Data generated from "multiplicative" process (populations, dollars)

Don't solve logs yourself; only need to `log(x)`

Undoing logs: `exp(log_x)`

